Supplementary Information: Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments

### 2025-06-25 doc/manuscript_hkk_drought_sensitivity.Rmd
update with word count and submission format

**Number of Words in Abstract:** 150  
**Number of Words in Main Text:** 5016  
<!-- Words to subtract: 2628-->
**Number of References:** 54  
**Number of Figures:** 5  
**Number of Tables:** 1  
**Number of Text boxes:** 0

**Corresponding author address:** anujank@si.edu

@@

# Abstract (150 words) {-}
An observed increase in the frequency and severity of droughts threatens forest health worldwide, with important implications for this critical carbon sink. 
Tree species adaptations to hot and dry conditions--for example, dry season deciduousness in tropical seasonal forests--might confer drought resistance, but such resistance may be variable across species, microenvironments, and drought events. 
We analysed dendrometer band measurements on 1820 trees across 30 species in a seasonally dry tropical forest in Western <!---Huai Kha Khaeng, ---->Thailand <!--using Bayesian regressions in a causal modelling framework---> to assess tree growth responses to three climatically distinct droughts. 
Species and individuals exhibited a wide range of growth responses within each drought, and these responses differed in intensity and affect among the drought events. 
Deciduous species were more sensitive to wet-season drought than evergreen species, and wetter microclimates amplified this difference. 
Exposed trees showed more sensitivity to drought, but  larger sized trees or trees in wetter topographic positions seemed to buffer growth declines. 
Heterogeneous drought responses of species and individuals indicate potential spatiotemporal insurance effects in diverse forests in the face of increased drought.

@@

Intense droughts are becoming more frequent, affecting the demography, diversity, and  carbon (C) cycling of forests globally [@ipcc_climate_2023]. 
However, drought impacts on tree woody growth and ecosystem C sequestration remain difficult to predict, particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019] but for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022]. 
For forests worldwide, studies have separately shown that drought characteristics, species traits, and microenvironment affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter understanding of forest responses to drought. 
Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, tree drought responses may be conserved, and therefore linearly predictable; i.e., species with low sensitivity to past droughts would have low sensitivity in any microhabitat and any drought. 
In contrast, if these factors interact, species and individuals that exhibit resistance to one drought may prove more vulnerable to a drought with different characteristics. 
The complexity and diversity of drought responses may act as a mechanism for maintaining species diversity and buffering ecosystem function even as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022;@naumann_global_2018].

@@

<!--PARAGRAPH ON DROUGHT CHARACTERISTICS-->
<!-- Sean::: This paragraph is a bit redundant and all over the place. 1) in the tropics, seasonal dry periods (often defined as months with <= 100 mm rainfall) can have complex influences on tree phenology, from leaf turnover to growth allocation. Droughts, which can exacerbate or extend a dry season, or reduce cloud cover ina rainy season, would be expected interact with seasonal phenology also in complex ways. [Then go into some details with citations on ?] -->
Drought is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the timing, severity and duration of climatic anomalies [@slette_how_2019] with potentially distinct ecological consequences.
<!-- ; we define it here as hot/dry conditions anomalously departing from climatic means during any time of year  -->
Although forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], droughts with different seasonal timing, duration, or intensity may have distinct effects on tree growth, especially in seasonal forests with diverse adaptive strategies to handle regular dry periods [@albert_cryptic_2019;@gao_dynamic_2018].
<!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
<!-- Sean::: don't think the above is true universally. Laura Boeschoten and Erick Esteban have separate papers well in prep showing some species demonstarating seasonal growth and others not -->
In tropical forests with dry seasons unfavorable to growth, tree drought responses may vary dramatically depending on the timing of drought relative to seasonal cycles of climate, photosynthesis, and woody growth [@garcía-cervigón_climate_2020]. <!---- agreed but see above ref I put in to [@albert_ecology_letters_cryptic_phenology]. It seems like these arguments are a bit redundant. Or meandering. ---->
<!--(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable eith anomalous hot, dry conditions relative to seasonal expectations.) -->
Analyses to date have found that interannual variation in tropical tree growth <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly sensitive to dry season precipitation [@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent a very limited set of tropical tree species and climates. 
It remains unknown how diverse assemblages of tropical tree species <!--in tropical dry forests -->respond to different types of droughts, including those that intensify or conflict with regular seasonal cycles<!-- of wet and dry seasons-->. 
<!--
Seasonally dry tropical forest climates are often close to thermal and water stress tolerance thresholds for many species therein, making these forests vulnerable to drought [@allen_will_2017] and leading to stand-level patterns of drought-associated water stress [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019]. <!--the last phrase of this sentence and how it relates to Bennett et al. is unclear to me.
Trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress.
These strategies might confer differential advantages under drought conditions. -->

@@

<!-- Sean::: mention rooting depth? [@chitra-tarak: root of the drought and chitra-tarak: new_phyt 2022] -->
Tree species adaptations to drought vary widely, often leading to classification of species as "drought tolerant" or "drought sensitive" [@guillemot_small_2022; @vico_how_2017a; @oliveira_linking_2021].
In tropical seasonal forests in particular, trees have evolved to optimize growth, reproduction, and survival (i.e., fitness) under dry seasons, with strategies ranging along a spectrum from hydraulic safety to efficiency, leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021].
Dry-season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid foliar water loss [@desouza_drought_2020]. 
Other strategies include deep roots (hydraulic safety, to access deeper water in the dry season), more negative leaf water potential at turgor loss point ($\pi_{tlp}$, hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), and large vessels (hydraulic efficiency)<!--but this may put them at risk of embolism--> [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021]. 
Under drought conditions, some of these strategies can lend drought resistance while others can amplify drought sensitivity, especially under severe droughts [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021]. 
<!-- This spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.  -->
<!--Some theoretical evidence that the relative advantage of evergreen and deciduous strategies might differ with changing precipitation regimes [@vico_how_2017a]--->
Moreover, besides being adaptively conserved, these strategies can also have developmental or ecophysiological controls.
For example, some species regulate negative $\pi_{tlp}$'s, leaf deciduousness, and root depth based on environmental cues, leading to intraspecific or interannual variation and potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; @hulshof_variation_2010].

@@

<!--HYPOTHESIS PARAGRAPH-->
We hypothesize that tree drought sensitivity is jointly and interactively shaped by drought characteristics, species adaptations, and microenvironments, resulting in variable drought responses of species and individuals across droughts, and of species across microhabitats. 
Specifically, we test predictions that:
i) both species and individuals respond differently to different droughts;
ii) deciduous species, which rely on wet seasons for growth, are particularly sensitive to wet season drought; and,
iii) exposed trees are more sensitive to drought, but water availability, either through larger topographic location or tree size (and corresponding rooting volume), can buffer growth declines.
We tested these hypotheses using  a 14-year record of dendrometer band measurements in a dry seasonal evergreen forest [@bunyavejchewin_seasonally_2011] in Thailand. 
During this period, the forest experienced three different dry periods, two associated with ENSO events, and all three differing with respect to timing and duration (Table 1). 
With data on 1820 trees across 30 species we built Bayesian regression models exploring how species characteristics and individual spatial context influenced growth responses to these droughts. 
We used a causal modelling framework to account for potential confounding relationships among the variables. 
This analysis adds important insight into the complexity of seasonal dry forest response to drought in species-rich tropical forests.

@@

```{r table, echo=FALSE, warning=FALSE, message=FALSE}
# figure 0-----------------------------

@@

Our study site was the Huai Kha Khaeng ForestGEO plot (15.6324 N, 99.217 E), hereafter "HKK", in the Huai Kha Khaeng Wildlife Sanctuary, Uthai Thani, Thailand.
This is a 50-hectare plot in dry seasonal evergreen forest, with mean annual precipitation of 1400 mm, mean annual temperature of 23.5 C and a strong dry season from November to April [@anderson-teixeira_ctfsforestgeo_2015].
The region is periodically affected by droughts associated with the *El Niño* Southern Oscillation [ENSO, @räsänen_spatial_2016].
The dominant canopy species are _Hopea odorata_, _Vatica harmandiana_, _Dipterocarpus alatus_ and _Tetrameles nudiflora_ and lower layers of the canopy are dominated by sub-canopy evergreen and deciduous species like _Miliusa horsfieldii_ and _Polyalthia viridis_. 
<!-- For analysis, we focused on 30 of the most abundant species in the plot.    -->
<!-- Analysis of long-term tree-ring data shows that interannual variation tree growth at HKK is influenced by temperature and rainfall [@vlam_temperature_2014].   -->

![***Figure `r fig_droughts`. Climatic characteristics of the focal drought years (red in 2010, brown in 2015) compared with the average across years*** 2008-2019 (grey dashed) and its standard deviation (grey shading) in the Huai Kha Khaeng ForestGEO plot, for four daily climate variables: dry days, precipitation, maximum temperature and VPD.  (a) 30-day running means. (b) Anomalies of the drought year running means from the averages, in units of standard deviations.  (c) Standardised Precipitation Evapotranspiration Index (SPEI) values for the drought years at four different time scales for each month for each drought year. A k-month SPEI integrates over k months ending in the month current month. Across all panels, light blue background represents the average wet season, from May to October. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/10](display/Fig1.png)

In HKK, metal dendrometer bands were installed in 2008 on 2353 trees across 152 species following the standard ForestGEO protocol [@muller-landau_metal_2008]. 
Selected were a size-stratified and spatially stratified random sample, with additional trees chosen to fill out sample sizes for selected species of interest.
Thus relative abudances among the selected trees largely reflected abundance variation in the plot as a whole, with more individuals banded from the abundant species (n=1 to 230 trees per species). 
On each tree, the dendrometer band was installed at a height of 10 cm above the regular point of measurement for the whole plot census meaning installation at 1.4 m on most trees, and at higher heights on buttressed trees with higher measurement points. 
Dendrobands were censused twice a year with the measurement periods typically centred in June and January. 
During each census, the size of the window opening along the band was measured using Mitutoyo digital callipers (precision = 0.01 mm) and the diameter of the tree at the dendrometer band was measured using standard DBH tape.

@@

Using the late wet/early dry dendrometer band censuses, we calculated annualised growth for each year and tree. 
<!-- We first conducted quality control on the dataset, removing measurements that appeared likely to be data entry errors (misplaced decimals), potential misidentified bands (old band numbers that restarted after stopping) or misidentified individuals (individuals with conflicting metadata across censuses). -->
We first conducted quality control on window size measurements (see supplementary methods) and combined them with the diameter measurement at installation $(d_{1})$ to calculate diameter at each dendroband census using standard equations that uses the chord across the measurement window to estimate the arc, assuming a circular cross section[@detto_correcting_2023].
We used a direct measurement of $d_{1}$ at the time of installation if it was available; if not, we excluded window size measurements until the first available diameter measurement. 
We calculated annualised diameter increments for each individual $i$ for each year $t$ by finding the difference in diameter from the previous year ($inc_{i,t} = d_{i,t} - d_{i,t-1}$) and adjusting for the number of days ($tdif$) between measurements ($ainc_{i, t} = inc_{i,t} * \frac{tdif}{365}$).

<!-- To ensure that the data did not smooth over multiple years, we excluded increment measures spanning missing values for $d_{t-1}$. -->
<!-- We excluded large measurement outliers (> 3 standard deviations from the mean increment across all observations).  -->
We conducted further quality control on increments (see supplementary methods for details) to assemble the final dataset. 
Diameter tape measurements themselves are not precise enough for calculating drought sensitivity; error for growth is twice measurement error = 2 mm. 
In comparison, median annual growth across trees in HKK is ~1.4 mm and calculating deviation from annual growth needs sub-mm precision. 
However, we used tape measurements to flag likely errors in the dendrometer dataset. 
<!-- We calculated increments from annual tape measurements made on each dendrobanded tree at each census and removed dendroband measurements that had low agreement.  -->
Finally, we excluded trees with negative increments over the whole timeseries, and included only species that had at least 10 individuals, resulting in a final dataset of 1820 individuals of 30 species.

![***Figure `r fig_drought_responses`: Growth increment timeseries and drought responses*** across trees and species in the Huai Kha Khaeng ForestGEO plot. a) Median of annualised growth increments across ten dominant species (colours) and all trees (black) from 2009 to 2022. b) Distribution of drought sensitivity of tree growth across all individuals in the two drought years (see main text for equation). For comments - https://github.com/forestgeo/growth-precip-thailand/issues/11](display/Fig2_alt.png)

@@

The droughts analysed had distinct climatic and temporal characteristics, with 2010 primarily a dry-season drought, 2015 a wet-season drought and 2020 showing mixed characteristics (Table 1, Fig `r fig_droughts`).
The 2010 drought showed 30-day-mean meteorological anomalies between February and May including lower precipitation (min anomaly **<-1** SD), higher number of dry days (anomaly **~1.5** SD), higher maximum temperature (anomaly **>2** SD), and higher VPD (anomaly **>2** SD), including in the wet season.
<!-- Multi-month SPEI values reached their minima in May, after which unusually wet conditions led to a fairly rapid recovery to normal conditions. -->
1- to 6-month SPEI values reached minima between -1.5 and -2, indicative of severe drought, while the 12-month values remained above -1, indicative of mild drought. 
In 2015, pronounced meteorological anomalies on the 30-day scale occurred during the wet season, including precipitation (anomaly **<-1.5** SD), number of dry days (max anomaly **>2** SD), maximum monthly temperatures (anomaly **+1.5** SD), and monthly maximum VPD (max anomaly **+2** SD).
<!-- 1-month SPEI was <-1 in May and June, reaching -2 in May, the lowest in the the entire dendroband census period.  -->
<!-- Multi-month SPEI values reached their minima between June and August and remained negative into the subsequent dry season. -->
1- to 12-month SPEI values crossed minima -2, lowest in the dendroband census period and indicative of severe drought. 
In 2020, meteorological anomalies were pronounced both in the dry and wet seasons, including maximum temperature (**>2** SD) and VPD (**>2** SD). 
3- and 6-month SPEI values reached minima ~2 in February, indicative of borderline severe drought.
12-month SPEI showed severe drought conditions from January to April, and moderate drought from May to September, spanning both dry and wet seasons.

@@

![***Figure `r fig_species`: Variation in sensitivity among species*** in the Huai Kha Khaeng ForestGEO plot during droughts in 2010, 2015 and 2020. a) Relationship of the mean species sensitivities from model fits of an intercept-only model, with mean species deciduousness values. Line shown for significant correlation. b) Predicted sensitivity values across the observed range of deciduousness and topographic wetness index from a model with TWI, deciduosness and their interaction across all trees modelled for each drought year separately. For comments: https://github.com/forestgeo/growth-precip-thailand/issues/12](display/Fig3.png)

@@

![***Figure `r fig_microenvs`: Drivers of variation in sensitivity among individuals***. All panels represent model results from models (one for each year) predicting sensitivity with microenvironment conditions with species random effect on all slopes. a) Coefficient plots showing median effects and 90% CI for wetness, exposure and DBH of tree. b) Predicted relationship of drought sensitivity with topographic wetness index across all species derived from model predictions. Black line represents overall relationship (panel a); coloured lines for each species reflects deciduousness values. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/13](display/Fig4.png)

@@

![***Figure `r fig_dag`: Effects of size, exposure and water availability*** from combined models. Hypothesised relationships between microenvironment variables and drought sensitivity with their modelled effects in the HKK ForestGEO plot in the droughts of a) 2010, b) 2015 and c) 2020. Solid lines represent relationships where 90% credible intervals do not overlap 0, blue and red lines represent positive and negative effects respectively and line thickness is scaled to the effect size. d) Modelled sensitivity across crown exposure classes in 2010, 2015 and 2020. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/14](display/Fig5.png)

@@

>>>>

**Number of Words in Abstract:** 144  
**Number of Words in Main Text:** 5019  
<!-- Words to subtract: 2628-->
**Number of References:** 54  
**Number of Figures:** 5  
**Number of Tables:** 1  
**Number of Text boxes:** 0

**Corresponding author address:**  
Krishna Anujan,
ForestGEO Ecosystems and Climate Lab,  
Smithsonian's National Zoo and Conservation Biology Institute,  
Front Royal, VA 22630  
USA  
Phone: 
Email: anujank@si.edu

@@

# Abstract (150 words) {-}
Increased frequency and severity of droughts threatens forest health worldwide, with important implications for this critical carbon sink. 
Tree species adaptations to hot and dry conditions--e.g., dry-season deciduousness in tropical seasonal forests--might confer drought resistance, but such resistance may be variable across species, microenvironments, and drought events. 
We analysed dendrometer band measurements on 1820 trees across 30 species in a seasonally dry tropical forest in Western Thailand to assess tree growth responses to three climatically distinct droughts. 
Species and individuals exhibited a wide range of growth responses within each drought, and differences in response intensity and affect among the drought events. 
Deciduous and evergreen species were more sensitive to wet- and dry-season drought respectively, and wetter microclimates amplified this difference. 
<!-- Exposed trees showed more sensitivity to drought, but  larger sized trees or trees in wetter topographic positions seemed to buffer growth declines.  -->
Heterogeneous drought responses of species and individuals indicate potential spatiotemporal insurance effects in diverse forests in the face of increased drought.

@@

Intense droughts are becoming more frequent, affecting the demography, diversity, and  carbon (C) cycling of forests globally [@ipcc_climate_2023]. 
However, drought impacts on tree woody growth and ecosystem C sequestration remain difficult to predict, particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019] but for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022]. 
For forests worldwide, studies have separately shown that drought characteristics, species traits, and microenvironment affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter understanding of forest responses to drought. 
Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, tree drought responses may be conserved, and linearly predictable; i.e., species with low sensitivity to past droughts would have low sensitivity in any microhabitat and any drought. 
In contrast, if these factors interact, species and individuals that exhibit resistance to one drought may prove vulnerable to a drought with different characteristics. 
The complexity and diversity of drought responses may act as a mechanism for maintaining species diversity and buffering ecosystem function even as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022;@naumann_global_2018].

@@

<!--PARAGRAPH ON DROUGHT CHARACTERISTICS-->
<!-- Sean::: This paragraph is a bit redundant and all over the place. 1) in the tropics, seasonal dry periods (often defined as months with <= 100 mm rainfall) can have complex influences on tree phenology, from leaf turnover to growth allocation. Droughts, which can exacerbate or extend a dry season, or reduce cloud cover ina rainy season, would be expected interact with seasonal phenology also in complex ways. [Then go into some details with citations on ?] -->
Drought is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the timing, severity and duration of climatic anomalies [@slette_how_2019] with potentially distinct ecological consequences.
<!-- ; we define it here as hot/dry conditions anomalously departing from climatic means during any time of year  -->
Although forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], droughts with different seasonal timing, duration, or intensity may have distinct effects on tree growth, especially in seasonal forests with diverse adaptive strategies to handle regular dry periods [@albert_cryptic_2019;@gao_dynamic_2018].
<!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
<!-- Sean::: don't think the above is true universally. Laura Boeschoten and Erick Esteban have separate papers well in prep showing some species demonstarating seasonal growth and others not -->
In tropical forests with dry-seasons unfavorable to growth, tree drought responses may vary dramatically depending on the timing of drought relative to seasonal cycles of climate, photosynthesis, and woody growth [@garcía-cervigón_climate_2020]. <!---- agreed but see above ref I put in to [@albert_ecology_letters_cryptic_phenology]. It seems like these arguments are a bit redundant. Or meandering. ---->
<!--(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable eith anomalous hot, dry conditions relative to seasonal expectations.) -->
Analyses to date have found that interannual variation in tropical tree growth <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly sensitive to dry-season precipitation [@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent a very limited set of tropical tree species and climates. 
It remains unknown how diverse assemblages of tropical tree species <!--in tropical dry forests -->respond to different types of droughts, including those that intensify or conflict with regular seasonal cycles<!-- of wet and dry seasons-->. 
<!--
Seasonally dry tropical forest climates are often close to thermal and water stress tolerance thresholds for many species therein, making these forests vulnerable to drought [@allen_will_2017] and leading to stand-level patterns of drought-associated water stress [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019]. <!--the last phrase of this sentence and how it relates to Bennett et al. is unclear to me.
Trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress.
These strategies might confer differential advantages under drought conditions. -->

@@

<!-- Sean::: mention rooting depth? [@chitra-tarak: root of the drought and chitra-tarak: new_phyt 2022] -->
Tree species adaptations to drought vary widely, often leading to classification of species as "drought tolerant" or "drought sensitive" [@guillemot_small_2022; @vico_how_2017a; @oliveira_linking_2021].
In tropical seasonal forests in particular, trees have evolved to optimize growth, reproduction, and survival (i.e., fitness) under dry seasons, with strategies ranging along a spectrum from hydraulic safety to efficiency, leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021].
Dry-season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry-season to avoid foliar water loss [@desouza_drought_2020]. 
Other strategies include deep roots (hydraulic safety, to access deeper water in the dry-season), more negative leaf water potential at turgor loss point ($\pi_{tlp}$, hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), and large vessels (hydraulic efficiency)<!--but this may put them at risk of embolism--> [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021]. 
Under drought conditions, some of these strategies can lend drought resistance while others can amplify drought sensitivity, especially under severe droughts [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021]. 
<!-- This spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.  -->
<!--Some theoretical evidence that the relative advantage of evergreen and deciduous strategies might differ with changing precipitation regimes [@vico_how_2017a]--->
Moreover, besides being adaptively conserved, these strategies can also have developmental or ecophysiological controls.
For example, some species regulate negative $\pi_{tlp}$'s, leaf deciduousness, and root depth based on environmental cues, leading to intraspecific or interannual variation and potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; @hulshof_variation_2010].

@@

<!--HYPOTHESIS PARAGRAPH-->
We hypothesize that tree drought sensitivity is jointly and interactively shaped by drought characteristics, species adaptations, and microenvironments, resulting in variable drought responses of species and individuals across droughts, and of species across microhabitats. 
Specifically, we test predictions that:
i) both species and individuals respond differently to different droughts;
ii) deciduous species, which rely on wet-seasons for growth, are particularly sensitive to wet-season drought; and,
iii) exposed trees are more sensitive to drought, but water availability, either through larger topographic location or tree size (and corresponding rooting volume), can buffer growth declines.
We tested these hypotheses using  a 14-year record of dendrometer band measurements in a dry seasonal evergreen forest [@bunyavejchewin_seasonally_2011] in Thailand. 
During this period, the forest experienced three different dry periods, two associated with ENSO events, and all three differing with respect to timing and duration (Table 1). 
With data on 1820 trees across 30 species we built Bayesian regression models exploring how species characteristics and individual spatial context influenced growth responses to these droughts. 
We used a causal modelling framework to account for potential confounding relationships among the variables. 
This analysis adds important insight into the complexity of seasonal dry forest response to drought in species-rich tropical forests.

@@

```{r table, eval=F, echo=FALSE, warning=FALSE, message=FALSE}
# figure 0-----------------------------

@@

Our study site was the Huai Kha Khaeng ForestGEO plot (15.6324 N, 99.217 E), hereafter "HKK", in the Huai Kha Khaeng Wildlife Sanctuary, Uthai Thani, Thailand.
This is a 50-hectare plot in dry seasonal evergreen forest, with mean annual precipitation of 1400 mm, mean annual temperature of 23.5 C and a strong dry-season from November to April [@anderson-teixeira_ctfsforestgeo_2015].
The region is periodically affected by droughts associated with the *El Niño* Southern Oscillation [ENSO, @räsänen_spatial_2016].
The dominant canopy species are _Hopea odorata_, _Vatica harmandiana_, _Dipterocarpus alatus_ and _Tetrameles nudiflora_ and lower layers of the canopy are dominated by sub-canopy evergreen and deciduous species like _Miliusa horsfieldii_ and _Polyalthia viridis_. 
<!-- For analysis, we focused on 30 of the most abundant species in the plot.    -->
<!-- Analysis of long-term tree-ring data shows that interannual variation tree growth at HKK is influenced by temperature and rainfall [@vlam_temperature_2014].   -->

<!-- ![***Figure `r fig_droughts`. Climatic characteristics of the focal drought years (red in 2010, brown in 2015) compared with the average across years*** 2008-2019 (grey dashed) and its standard deviation (grey shading) in the Huai Kha Khaeng ForestGEO plot, for four daily climate variables: dry days, precipitation, maximum temperature and VPD.  (a) 30-day running means. (b) Anomalies of the drought year running means from the averages, in units of standard deviations.  (c) Standardised Precipitation Evapotranspiration Index (SPEI) values for the drought years at four different time scales for each month for each drought year. A k-month SPEI integrates over k months ending in the month current month. Across all panels, light blue background represents the average wet season, from May to October. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/10](display/Fig1.png) -->

In HKK, metal dendrometer bands were installed in 2008 on 2353 trees across 152 species following standard ForestGEO protocol [@muller-landau_metal_2008]. 
Selected were a size-stratified and spatially-stratified random sample, with additional trees chosen to fill out sample sizes for selected species of interest.
Thus relative abudances among the selected trees largely reflected abundance variation in the plot as a whole, with more individuals banded from the abundant species (n=1 to 230 trees per species). 
On each tree, the dendrometer band was installed at a height of 10 cm above the regular point of measurement for the whole-plot census meaning installation at 1.4 m on most trees, and at higher heights on buttressed trees with higher measurement points. 
Dendrobands were censused twice a year with the measurement periods typically centred in June and January. 
During each census, the size of the window opening along the band was measured using Mitutoyo digital callipers (precision = 0.01 mm) and the diameter of the tree at the dendrometer band was measured using standard DBH tape.

@@

Using the late wet/early dry dendrometer band censuses, we calculated annualised growth for each year and tree. 
<!-- We first conducted quality control on the dataset, removing measurements that appeared likely to be data entry errors (misplaced decimals), potential misidentified bands (old band numbers that restarted after stopping) or misidentified individuals (individuals with conflicting metadata across censuses). -->
We first conducted quality control on window size measurements (see supplementary methods) and combined them with the diameter measurement at installation $(d_{1})$ to calculate diameter at each dendroband census using standard equations that uses the chord across the measurement window to estimate the arc[@detto_correcting_2023].
We used a direct measurement of $d_{1}$ at the time of installation if it was available; if not, we excluded window size measurements until the first available diameter measurement. 
We calculated annualised diameter increments for each individual $i$ for each year $t$ by finding the difference in diameter from the previous year ($inc_{i,t} = d_{i,t} - d_{i,t-1}$) and adjusting for the number of days ($tdif$) between measurements ($ainc_{i, t} = inc_{i,t} * \frac{tdif}{365}$).

<!-- To ensure that the data did not smooth over multiple years, we excluded increment measures spanning missing values for $d_{t-1}$. -->
<!-- We excluded large measurement outliers (> 3 standard deviations from the mean increment across all observations).  -->
We conducted further quality control on increments (see supplementary methods for details) to assemble the final dataset. 
Diameter tape measurements are not precise enough for calculating drought sensitivity. 
Error for growth is twice measurement error = 2 mm; in comparison, median annual growth in HKK is ~1.4 mm and calculating deviation from annual growth needs sub-mm precision. 
However, we used tape measurements to flag likely errors in the dendrometer dataset. 
<!-- We calculated increments from annual tape measurements made on each dendrobanded tree at each census and removed dendroband measurements that had low agreement.  -->
Finally, we excluded trees with negative increments over the whole timeseries, and included only species that had at least 10 individuals, resulting in a final dataset of 1820 individuals of 30 species.

<!-- ![***Figure `r fig_drought_responses`: Growth increment timeseries and drought responses*** across trees and species in the Huai Kha Khaeng ForestGEO plot. a) Median of annualised growth increments across ten dominant species (colours) and all trees (black) from 2009 to 2022. b) Distribution of drought sensitivity of tree growth across all individuals in the two drought years (see main text for equation). For comments - https://github.com/forestgeo/growth-precip-thailand/issues/11](display/Fig2_alt.png) -->

@@

The droughts analysed had distinct climatic and temporal characteristics, with 2010 primarily a dry-season drought, 2015 a wet-season drought and 2020 showing mixed characteristics (Table 1, Fig `r fig_droughts`).
The 2010 drought showed 30-day-mean meteorological anomalies between February and May including lower precipitation (min anomaly **<-1** SD), higher number of dry days (anomaly **~1.5** SD), higher maximum temperature (anomaly **>2** SD), and higher VPD (anomaly **>2** SD), including in the wet-season.
<!-- Multi-month SPEI values reached their minima in May, after which unusually wet conditions led to a fairly rapid recovery to normal conditions. -->
1- to 6-month SPEI values reached minima between -1.5 and -2, indicative of severe drought, while the 12-month values remained above -1, indicative of mild drought. 
In 2015, pronounced meteorological anomalies on the 30-day scale occurred during the wet-season, including precipitation (anomaly **<-1.5** SD), number of dry days (max anomaly **>2** SD), maximum monthly temperatures (anomaly **+1.5** SD), and monthly maximum VPD (max anomaly **+2** SD).
<!-- 1-month SPEI was <-1 in May and June, reaching -2 in May, the lowest in the the entire dendroband census period.  -->
<!-- Multi-month SPEI values reached their minima between June and August and remained negative into the subsequent dry season. -->
1- to 12-month SPEI values crossed minima -2, lowest in the dendroband census period and indicative of severe drought. 
In 2020, meteorological anomalies were pronounced both in the dry- and wet-seasons, including maximum temperature (**>2** SD) and VPD (**>2** SD). 
3- and 6-month SPEI values reached minima ~2 in February, indicative of borderline severe drought.
12-month SPEI showed severe drought conditions from January to April, and moderate drought from May to September, spanning both dry- and wet-seasons.

@@

<!-- ![***Figure `r fig_species`: Variation in sensitivity among species*** in the Huai Kha Khaeng ForestGEO plot during droughts in 2010, 2015 and 2020. a) Relationship of the mean species sensitivities from model fits of an intercept-only model, with mean species deciduousness values. Line shown for significant correlation. b) Predicted sensitivity values across the observed range of deciduousness and topographic wetness index from a model with TWI, deciduosness and their interaction across all trees modelled for each drought year separately. For comments: https://github.com/forestgeo/growth-precip-thailand/issues/12](display/Fig3.png) -->

@@

<!-- ![***Figure `r fig_microenvs`: Drivers of variation in sensitivity among individuals***. All panels represent model results from models (one for each year) predicting sensitivity with microenvironment conditions with species random effect on all slopes. a) Coefficient plots showing median effects and 90% CI for wetness, exposure and DBH of tree. b) Predicted relationship of drought sensitivity with topographic wetness index across all species derived from model predictions. Black line represents overall relationship (panel a); coloured lines for each species reflects deciduousness values. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/13](display/Fig4.png) -->

@@

<!-- ![***Figure `r fig_dag`: Effects of size, exposure and water availability*** from combined models. Hypothesised relationships between microenvironment variables and drought sensitivity with their modelled effects in the HKK ForestGEO plot in the droughts of a) 2010, b) 2015 and c) 2020. Solid lines represent relationships where 90% credible intervals do not overlap 0, blue and red lines represent positive and negative effects respectively and line thickness is scaled to the effect size. d) Modelled sensitivity across crown exposure classes in 2010, 2015 and 2020. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/14](display/Fig5.png) -->

@@

<div id="refs"></div>

# Table Caption {-}

**Table 1. Summary table of hypotheses tested, effects of variables on drought sensitivity and corresponding evidence.** For each variable x year combination, "+" indicates a positive effect, "-" indicates a negative effect and "n.s." indicates that the effect was not statistically significant.
